Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2019-12-30 manuscript/abstract.tex
Copyedit main and abstract text
Correct spelling and clarify wording.

Seasonal influenza virus A/H3N2 is a major cause of illness and death globally.
While vaccination is the most effective way to prevent infection, rapid accumulation of mutations in the surface protein hemagglutinin (HA) allows viruses to escape previous adaptive immunity due to vaccination or natural infection.
This antigenic drift requires vaccines to be updated regularly.
Due to a nearly one-year lag between the selection of vaccine strains and deployment of the vaccine to the public, effective vaccine strain selection requires predictions about future influenza populations.
Historically, experts predicted successful future viruses through manual inspection of hemagglutination inhibition (HI) assays, an experimental measure of antigenic drift.
Modern strain selection is increasingly informed by modeling of influenza fitness using HA sequences to estimate antigenic drift, mutational load, and recent population growth.
The models rely entirely on genotypic information without accounting for the gold standard phenotypic measures of antigenic drift provided by HI assays.
We developed a novel open source forecasting framework based on these original fitness models that explicitly integrates genotypic and phenotypic measures of antigenic drift as well as modern experimental measurements of mutational load from deep mutational scanning (DMS) experiments.
In contrast with previous work, our models minimize the distance between observed and estimated future populations using the Earth Mover's Distance metric instead of minimizing errors in frequencies of phylogenetic clades.
We show that this integrated approach to forecasting can accurately estimate the sequence compositions of simulated and natural populations.
These forecasts capture dynamics of influenza clades and enable the identification of optimal individual strains for vaccine composition.
Using this framework and a modern machine learning approach to model testing, we found that phenotypic measures of antigenic drift were more consistently predictive of future influenza populations than sequence-only measures while the opposite was true for measures of mutational load.
Importantly, the combination of antigenic drift and mutational load was the most predictive of future success.
We provide real-time forecasts of seasonal influenza A/H3N2 populations based on our combined model of antigenic drift and mutational load as part of Nextstrain.
As our open source framework allows definition of fitness metrics per sample using tidy data frames and does not require phylogenetic clades for forecasts, we hope it paves the way for future forecasting efforts with pathogens that cannot be readily analyzed by standard phylogenetic methods including recombinant viruses, bacteria, and fungi.

>>>>

Seasonal influenza virus A/H3N2 is a major cause of illness and death globally.
While vaccination is the most effective way to prevent infection, rapid accumulation of mutations in the surface protein hemagglutinin (HA) allows viruses to escape previous adaptive immunity due to vaccination or natural infection.
This antigenic drift requires vaccines to be updated regularly.
Due to a nearly one-year lag between the selection of vaccine strains and deployment of the vaccine to the public, effective vaccine strain selection requires predictions about future influenza populations.
Historically, experts predicted successful future viruses through manual inspection of hemagglutination inhibition (HI) assays, an experimental measure of antigenic drift.
Modern strain selection is increasingly informed by modeling of influenza fitness using HA sequences to estimate antigenic drift, functional constraint, and recent population growth.
The models rely entirely on genotypic information without accounting for the gold standard phenotypic measures of antigenic drift provided by HI assays.
We developed a novel open source forecasting framework based on these original fitness models that explicitly integrates genotypic and phenotypic measures of antigenic drift as well as modern experimental measurements of functional constraint from deep mutational scanning (DMS) experiments.
In contrast with previous work, our models minimize the distance between observed and estimated future populations using the Earth Mover's Distance metric instead of minimizing errors between observed and estimated phylogenetic clade frequencies.
We show that this integrated approach to forecasting can accurately estimate the sequence compositions of simulated and natural populations.
These forecasts capture dynamics of influenza clades and enable the identification of optimal individual strains for vaccine composition.
Using this framework and a modern machine learning approach to model testing, we found that phenotypic measures of antigenic drift were more consistently predictive of future influenza populations than sequence-only measures while the opposite was true for measures of functional constraint.
Importantly, the combination of antigenic drift and mutational load was the most predictive of future success.
We provide real-time forecasts of seasonal influenza A/H3N2 populations based on our combined model of antigenic drift and mutational load as part of Nextstrain.
As our open source framework allows definition of fitness metrics per sample using tidy data frames and does not require phylogenetic clades for forecasts, we hope it paves the way for future forecasting efforts with pathogens that cannot be readily analyzed by standard phylogenetic methods including recombinant viruses, bacteria, and fungi.

### 2019-12-30 manuscript/main.tex
Copyedit main and abstract text
Correct spelling and clarify wording.

\begin{figure*}[ht!]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations-controls.png}
  \caption{
    Model coefficients and distances between projected and observed future populations as measured in amino acids (AAs) for simulated populations.
    A) Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
    For model testing, coefficients were fixed to their mean values from training/validation and applied to out-of-sample test data (open circles, N=18).
    B) Distances between projected and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
    The mean $\pm$ standard deviation of distances per validation timepoint are shown in the top-left of each panel.
    Corresponding values per test timepoint are in the top-right.
    For reference, the naive model's distance to the future is also shown per validation and test timepoint in grey.
    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 9.07 $\pm$ 1.71 AAs for test timepoints.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}
  \end{center}
\end{figure*}

@@

\begin{figure*}[p]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.png}
  \caption{
  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model for 18 timepoints shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  The estimated best strain was in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure*}

@@

As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$, $p < 0.001$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
These results confirm that our approach of minimizing the distance between yearly populations can simulataneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
These results also highlight a well-established yet critical point that forecasts on truly out-of-sample data are less accurate than those based on training and validation data.

@@

\begin{figure*}[p]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-natural-populations.png}
  \caption{
  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity and mutational load.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future for 8 timepoints.
  The estimated best strain was in the top 20th percentile of observed closest strains for 88\% of timepoints.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (grey shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  }
  \label{fig:test_of_best_model_for_natural_populations}
  \end{center}
\end{figure*}

@@

We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza A/H3N2 by estimating the sequence composition of future populations.
A key innovation of this framework is its ability to directly compare viral populations between seasons using the Earth Mover's Distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural A/H3N2 populations without relying on clades as model targets.
We have further introduced novel fitness metrics based on experimental measurements of antigenic drift and functional constraint.
We demonstrated how these metrics can be effectively integrated with previously published sequence-only metrics.
We have added this framework as a component of seasonal influenza analyses on nextstrain.org where it provides real-time forecasts for influenza researchers, decision makers, and the public.

@@

Our evaluation of models by time-series cross-validation and true out-of-sample forecasts revealed substantial potential for model overfitting.
We observed overfitting to both specific genetic backgrounds and general historical contexts.
A clear example of the former was the poor performance of our DMS-based fitness metric compared to a simpler mutational load metric.
Although the DMS experiments provided detailed estimates of which amino acids were preferred at which positions in HA, these measurements were specific to a single strain, A/Perth/16/2009 \cite{Lee2018}.
When we applied these measurements to predict the success of global populations, they were less informative on average than the naive model.
To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple A/H3N2 strains from distinct historical contexts.
We anticipate that these measurements could be used to define and continually update a modern set of sites contributing to mutational load in natural populations.
This set of sites could replace the statically defined set of ``non-epitope'' sites we use to estimate mutational load here.

We observed two different cases of overfitting to general historical contexts.
In the first case, the model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
To avoid such overfitting in the future, models will need fitness metrics that can represent the evolutionary contributions of neuraminidase, specifically, or of reassortment, generally.

@@

By forecasting the composition of future A/H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AA per year for validation data and 0.66 AA per year for test data.
However, these results are consistent with the observed dynamics of A/H3N2.
First, the one-year forecast horizon is a fraction of the average coalescence time for A/H3N2 populations of about 3--8 years \cite{Rambaut:2008ew}.
Hence, we expect the diversity of circulating strains to persist between seasons.
Second, the average substitution rate for A/H3N2's HA is 0.0043 substitutions per site per year or an average accumulation of 2.4 AAs per year \cite{}.
This accumulation of substitutions contributes to the distance between annual populations observed by the naive model.
In this context, our model gains of 0.96 and 0.66 AAs per year correspond to an explanation of 40\% and 28\% of the expected additional distance between annual populations, respectively.

There are still several clear opportunities to improve forecasts.
Integration of more recent experimental data may improve estimates of antigenic drift.
Influenza collaborating centers now routinely measure antigenic drift by focus reduction assays (FRAs), as an alternative to hemagglutination inhibition (HI) assays that is less affected by glycosylation motifs that alter binding efficacy \cite{Zost2017}.
While there are not enough historical FRA measurements to extensively validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
Further experiments will be required to fully understand how HI and FRA measurements of the same pairs of strains correlate with each other.
In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape A/H3N2 strains must navigate to be globally successful.

@@

To account for uncertainty in collection date and sampling error, we applied a kernel density estimation (KDE) approach to calculate global sample frequencies.
Specifically, we constructed a Gaussian kernel for each sample with the mean at the reported collection date and a variance (or KDE bandwidth) of two months.
The bandwidth was identified by cross-validation, as described above.
This bandwidth also roughly corresponds to the median lag time between sample collection and submission to the GISAID database.
We estimated the frequency of each sample at each timepoint by calculating the probabilitiy density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
We implemented this frequency estimation logic in the augur frequencies command.

@@

For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of samples at time $t$ by their estimated fitnesses.
We measured distance between populations using Earth Mover's Distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
This solution requires both a ``ground distance'' between pairs fo samples from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.

>>>>

\begin{figure*}[ht!]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations-controls.png}
  \caption{
    Model coefficients and distances between projected and observed future populations as measured in amino acids (AAs) for simulated populations.
    A) Coefficients are shown per validation timepoint (solid circles, N=33) with the mean $\pm$ standard deviation in the top-left corner.
    For model testing, coefficients were fixed to their mean values from training/validation and applied to out-of-sample test data (open circles, N=18).
    B) Distances between projected and observed populations are shown per validation timepoint (solid black circles) or test timepoint (open black circles).
    The mean $\pm$ standard deviation of distances per validation timepoint are shown in the top-left of each panel.
    Corresponding values per test timepoint are in the top-right.
    For reference, the naive model's distance to the future is also shown per validation and test timepoint in gray.
    The naive model's distance to the future was 8.97 $\pm$ 1.35 AAs for validation timepoints and 9.07 $\pm$ 1.71 AAs for test timepoints.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}
  \end{center}
\end{figure*}

@@

\begin{figure*}[p]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-simulated-populations.png}
  \caption{
  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model for 18 timepoints shows how often the model makes a good choice when forced to select a single representative strain for the future population.
  The estimated best strain was in the top 20th percentile of observed closest strains for 89\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure*}

@@

As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
While observed and estimated growth rates correlated more weakly for test data (Pearson's $R^2 = 0.14$, $p < 0.001$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}A).
Correspondingly, we observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}C).
The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}B).
Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, $p < 0.001$, Supplemental Fig.~\ref{fig:test_of_best_model_for_simulated_populations}D).
These results confirm that our approach of minimizing the distance between yearly populations can simultaneously capture clade-level dynamics of these populations and identify optimal individual strains as vaccine candidates.
These results also highlight a well-established yet critical point that forecasts on truly out-of-sample data are less accurate than those based on training and validation data.

@@

\begin{figure*}[p]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/test-of-best-model-for-natural-populations.png}
  \caption{
  Test of best model for natural populations of A/H3N2 viruses, the composite model of HI cross-immunity and mutational load.
  A) The correlation of estimated and observed clade growth rates shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future for 8 timepoints.
  The estimated best strain was in the top 20th percentile of observed closest strains for 88\% of timepoints.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  }
  \label{fig:test_of_best_model_for_natural_populations}
  \end{center}
\end{figure*}

@@

We have developed and rigorously tested a novel, open source framework for forecasting the long-term evolution of seasonal influenza A/H3N2 by estimating the sequence composition of future populations.
A key innovation of this framework is its ability to directly compare viral populations between seasons using the Earth Mover's Distance metric \cite{Rubner1998} and eliminate unavoidably stochastic clade definitions from phylogenies.
The best models from this framework still effectively capture clade dynamics and accurately identify optimal vaccine candidates from simulated and natural A/H3N2 populations without relying on clades as model targets.
We have further introduced novel fitness metrics based on experimental measurements of antigenic drift and functional constraint.
We demonstrated that the integration of these phenotypic metrics with previously published sequence-only metrics produces more accurate forecasts than sequence-only models.
We have added this framework as a component of seasonal influenza analyses on nextstrain.org where it provides real-time forecasts for influenza researchers, decision makers, and the public.

@@

Our evaluation of models by time-series cross-validation and true out-of-sample forecasts revealed substantial potential for model overfitting.
We observed overfitting to both specific genetic backgrounds and general historical contexts.
A clear example of the former was the poor performance of our DMS-based fitness metric compared to a simpler mutational load metric.
Although the DMS experiments provided detailed estimates of which amino acids were preferred at which positions in HA, these measurements were specific to a single strain, A/Perth/16/2009 \cite{Lee2018}.
When we applied these measurements to predict the success of global populations, they were less informative on average than the naive model.
To benefit from the more comprehensive fitness costs measured by DMS data, future models will need to synthesize DMS measurements across multiple A/H3N2 strains from distinct genetic contexts.
We anticipate that these measurements could be used to define and continually update a modern set of sites contributing to mutational load in natural populations.
This set of sites could replace the statically defined set of ``non-epitope'' sites we use to estimate mutational load here.

We observed two different cases of overfitting to general historical contexts.
In the first case, the model with the best performance across our validation data -- mutational load and LBI -- was also one of the worst models across our test data.
Specifically, we found that this model failed to predict the sudden decline of a dominant reassortant clade, A2/re, in 2019.
Despite this model's excellent performance historically, it was unable to account for rare yet important events such as reassortment.
To avoid such overfitting in the future, models will need fitness metrics that can balance the fitness benefits of evolution in other genes such as neuraminidase \cite{Chen:2018kp} with the costs of reassortment itself \cite{Villa:2017iw}.

@@

By forecasting the composition of future A/H3N2 populations with biologically-informed fitness metrics, our best models consistently outperformed a naive model (Table~\ref{table_natural_model_selection}).
While this performance confirms previously demonstrated potential for long-term influenza forecasting \cite{Luksza:2014hj}, the average gain from these models over the naive model appears low at 0.96 AAs per year for validation data and 0.66 AAs per year for test data.
However, these results are consistent with the observed dynamics of A/H3N2.
First, the one-year forecast horizon is a fraction of the average coalescence time for A/H3N2 populations of about 3--8 years \cite{Rambaut:2008ew}.
Hence, we expect the diversity of circulating strains to persist between seasons.
Second, the average substitution rate for A/H3N2's HA is 0.0043 substitutions per site per year or an average accumulation of 2.4 AAs per year \cite{}.
This accumulation of substitutions contributes to the distance between annual populations observed by the naive model.
In this context, our model gains of 0.96 and 0.66 AAs per year correspond to an explanation of 40\% and 28\% of the expected additional distance between annual populations, respectively.

Several clear opportunities to improve forecasts still remain.
Integration of more recent experimental data may improve estimates of antigenic drift.
Influenza collaborating centers now routinely measure antigenic drift by focus reduction assays (FRAs), as an alternative to hemagglutination inhibition (HI) assays that is less affected by glycosylation motifs that alter binding efficacy \cite{Zost2017}.
While there are not enough historical FRA measurements to extensively validate and test a FRA cross-immunity metric, it may be possible to reuse coefficients from the HI cross-immunity models with modern FRA data.
Further experiments will be required to fully understand how HI and FRA measurements of the same pairs of strains correlate with each other.
In addition to these FRA data based on ferret antisera, recent high-throughput antigenic escape assays with human sera promise to improve existing definitions of epitope sites \cite{Lee2019}.
These assays reveal the specific sites and residues that confer antigenic escape from polyclonal sera obtained from individual humans.
A sufficiently broad geographic and temporal sample of human sera with these assays could reveal consistent patterns of the immune landscape A/H3N2 strains must navigate to be globally successful.

@@

To account for uncertainty in collection date and sampling error, we applied a kernel density estimation (KDE) approach to calculate global sample frequencies.
Specifically, we constructed a Gaussian kernel for each sample with the mean at the reported collection date and a variance (or KDE bandwidth) of two months.
The bandwidth was identified by cross-validation, as described above.
This bandwidth also roughly corresponds to the median lag time between sample collection and submission to the GISAID database.
We estimated the frequency of each sample at each timepoint by calculating the probability density function of each KDE at that timepoint and normalizing the resulting values to sum to one.
We implemented this frequency estimation logic in the augur frequencies command.

@@

For a model based on any given combination of fitness metrics, we found the fitness coefficients that minimized the weighted Hamming distance between amino acid sequences from the observed future population at time $u = t + \Delta{t}$ and the estimated future population created by projecting frequencies of samples at time $t$ by their estimated fitnesses.
We measured distance between populations using Earth Mover's Distance (EMD), a metric commonly applied in machine learning to compare collections of pixels or words \cite{Rubner1998,Kusner2015}.
Solving for EMD identifies the minimum about of ``earth'' that must be moved from a source population to a sink population to make those populations as similar as possible.
This solution requires both a ``ground distance'' between pairs of samples from both populations and weights assigned to each sample that determine how much that sample contributes to the overall distance.
